MS Technical Paper: \\ Placement Algorithms for Heterogenous FPGAs

### 2025-03-26 latex/paper/paper.tex
more outlining

    In this paper, we recreate and experiment with established placement algorithms by leveraging the Xilinx's RapidWright, which is a semi-open-source API that offers backend access to Xilinx's industry-standard FPGA environment, Vivado. We implement a simulated annealing placer for Xilinx's 7-series FPGAs, with an emphasis on minimizing total wirelength while mitigating runtime. Our implementation is organized into three consecutive substages. The \textbf{prepacking} stage involves traversing a raw EDIF netlist to identify recurring cell patterns—such as CARRY chains, DSP cascades, and LUT-FF pairs—that are critical for efficient mapping and legalization. In the subsequent \textbf{packing} stage, these identified patterns, along with any remaining loose cells, are consolidated into SiteInst objects that encapsulate the FPGA’s discrete resource constraints and architectural nuances. Finally, the \textbf{placement} stage employs a simulated annealing (SA) algorithm to optimally assign SiteInst objects to physical sites, aiming to minimize total wirelength while adhering to the constraints of the 7-series architecture.

    Simulated annealing iteratively swaps placement objects, in our case, SiteInsts, guided by a cost function that decides which swaps should be accepted or rejected. Hill climbing is permitted by occasionally accepting moves that increase cost, in hope that such swaps may later lead to a better final solution. SA remains a popular approach in FPGA placement research due to its simplicity and robustness in handling the discrete architectural constraints of FPGA devices. While SA yields surprisingly good results given relatively simple rules, it is ultimately a heuristic and stochastic approach that explores the vast placement space by making random moves. Most of these moves will be rejected, meaning that SA must run many iterations, usually hundreds to thousands, to arrive at a desirable solution.

@@

\section{Xilinx 7-Series Architecture}
    Before any work can be done on a placer, one must consider what is being placed and upon what it is being placed. \\
    (Square peg square hole analogy) \\
    (Show the arch hierarchy). \\

@@

@@

    The RapidWright API offers a user-friendly toolkit to take pre-implemented designs, that is, designs that have already been placed and/or routed by Vivado, and make custom optimizations to fit complex design criteria. 
    RapidWright has several "BlockPlacers", a general Router, but no general Placer. 
    We use this API to implement a crude general placer, in our case, a simulated annealer. 
    (Show what parts of the FPGA toolchain that RapidWright already has and how our placer fits into that toolchain).

\section{Simulated Annealing}
    (Describe the algorithm, then in context of 7-Series arhictecture). \\

>>>>

    For these reasons, FPGA placement remains a critical research effort even today. 
    In this paper, we study and implement established placement methods by leveraging RapidWright, which is a semi-open-source API that offers backend access to Xilinx's industry-standard FPGA environment, Vivado. 
    We implement multiple variations of simulated annealing placers for Xilinx's 7-series FPGAs, with an emphasis on minimizing total wirelength while mitigating runtime. 
    Our implementation is organized into three consecutive substages. 
    The \textbf{prepacking} stage involves traversing a raw EDIF netlist to identify recurring cell patterns—such as CARRY chains, DSP cascades, and LUT-FF pairs—that are critical for efficient mapping and legalization. 
    In the subsequent \textbf{packing} stage, these identified patterns, along with any remaining loose cells, are consolidated into SiteInst objects that encapsulate the FPGA’s discrete resource constraints and architectural nuances. 
    Finally, the \textbf{placement} stage employs a simulated annealing (SA) algorithm to optimally assign SiteInst objects to physical sites, aiming to minimize total wirelength while adhering to the constraints of the 7-series architecture.

    Simulated annealing iteratively swaps placement objects guided by a cost function that decides which swaps should be accepted or rejected. 
    Hill climbing is permitted by occasionally accepting moves that increase cost, in hope that such swaps may later lead to a better final solution. 
    SA remains a popular approach in FPGA placement research due to its simplicity and robustness in handling the discrete architectural constraints of FPGA devices. 
    While SA yields surprisingly good results given relatively simple rules, it is ultimately a heuristic and stochastic approach that explores the vast placement space by making random moves. 
    Most of these moves will be rejected, meaning that SA must run many iterations, usually hundreds to thousands, to arrive at a desirable solution.

@@

\end{multicols}

\section{The FPGA Toolchain}
    Engineer design entry (Verilog, VHDL) \\
    Synthesis \\ 
    Placement \\ 
    Routing \\ 
    Bitstream generation \\

    (Square peg square hole analogy) \\
    (Show the architecture hierarchy). \\

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/hierarchy.png}
    \caption{Architecture Hierarchy of a Xilinx FPGA}\label{fig:hierarchy}
\end{figure}

\begin{multicols}{2}
\section{Xilinx 7-Series Architecture}
    Before any work can be done on a placer, one must understand the objects being placed and the structure of the medium in which those objects reside. 
    Figure~\ref{fig:hierarchy} illustrates the architectural hierarchy of a Xilinx FPGA. 
    At its most basic level, an FPGA is a vast physical array of replicated atomic components known as Basic Elements of Logic (BELs).
    These BELs, which include programmable lookup tables (LUTs), flip flops (FFs), and programmable interconnects, form the fundamental building blocks for custom digital circuit implementations on FPGAs.

    In modern FPGAs like the Xilinx 7-Series, additional BELs like Block RAM (BRAM) and Digital Processing Slices (DSPs) are also available, creating a more heterogeneous programmable logic fabric. 
    These dedicated resources allow designers to implement common digital structures, such as multipliers and memories, without having to recreate them from scratch using only LUTs and flip flops.

    The Xilinx architectural hierarchy organizes these BELs into increasingly abstract structures. 
    \textbf{BELs} are packaged into \textbf{Sites}, which are then embedded into \textbf{Tiles}, which are subsequently consolidated into \textbf{Clock Regions}. 
    In high-end Xilinx devices, Clock Regions may be further consolidated into \textbf{Super Logic Regions} (SLRs). 
    However, for the scope of this paper, we focus on Xilinx devices that have only a single SLR.

    Take for example, the xc7z020 chip, a low-end Xilinx device from the 7-Series family of FPGAs, commonly found on hobbyist boards like the Digilent Arty-Z7.

@@

    The RapidWright API offers a user-friendly toolkit to take pre-implemented designs, that is, designs that have already been placed and/or routed by Vivado, and make custom optimizations to fit complex design criteria. 
    RapidWright has several "BlockPlacers", a general Router, but no general Placer. 
    We use this API to implement a crude general placer, in our case, a simulated annealer. 
    (Show what parts of the FPGA toolchain that RapidWright already has and how our placer fits into that toolchain). 
    \subsection{Device Objects}
    \subsection{Design Objects}

\section{Placement Implementation}
    (Describe the algorithm, then in context of 7-Series arhictecture). \\
    \subsection{PrePacking}
        In the prepacking stage, we traverse the raw EDIF Netlist to identify recurring cell patterns. In particular, we will specifically look for CARRY chains, DSP cascades, and LUT-FF pairs.
        \subsection{CARRY Chains}
        \subsection{DSP Cascades}
        \subsection{LUT-FF Pairs}
            Identify unique CE-SR net pairs. All FF cells in a Site must share the same CE and SR nets.
    \subsection{Packing}
        In the packing stage, we take the identified cell clusters and pack them into SiteInst Design objects which target the Device Site objects.
        \subsection{CLB Sites}
            Can support LUT-FF pairs, loose LUTs, loose FFs, CARRY chains.
            CARRY chains span vertically across multiple Sites.
    \subsection{Placement}
